A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-08 latex/main.tex

The data collected during the COVID-19 pandemic as part of global
surveillance presents fundamental challenges to existing approaches
in two ways. Firstly, the sheer scale of data collected 
overwhelmed even approximate distance-matrix based methods.
At the peak in [month] 202X, there were over X thousand samples 
per day deposited in GISAID\cite{Shu2017}, with a total of 
over 17 million as of Sep 2025. % TODO check numbers
Regularly reinferring trees de-novo from such rapidly 
growing datasets is untenable.
The UShER\cite{Turakhia2021} method is based on adding 
samples to an existing tree using maximum parsimony, 
with periodic rearrangements to counter the inherent 
greediness of the approach\cite{DeMaio2023}.
The benefits of this simple model and the value of 
a comprehensive genealogy updated in real-time~\cite{McBroome2021}
were conclusively demonstrated, as the resource
quickly became an indispensible element 
of the pandemic response\cite{Hinrichs2024}.
The second challenge posed is that recombination is 
an important force in SARS-CoV-2 
evolution\cite{Jackson2021,VanInsberghe2021,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present in 
genealogy of the global population [cites??].
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{
Varabyou2021rapid,% Bolotie
Turakhia2022,%RIPPLES
Zhou2023virusrecom,% CovRecomb
Alfonsi2024,%RecombinHunt 
Li2024CovRecomb%CovRecomb 
},
this information is not incorporated into the phylogenies used 
in innumerable studies~\cite{McBroome2021,Hadfield2018},
and recombination continues to be treated as a 
nuisance factor.

@@

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts \cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango lineage naming system\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
In addition, we show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. By fully incorporating these and other novel
recombination events, the resulting data structure---representing
phylogenetic relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
% Dunno about this! Feels important to finish on a wider note, helping
% the reader understand the significance of all of this?
These resources
will help us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

@@

\subsection*{Recombination breakpoint intervals along the genome}

>>>>

The data collected during the COVID-19 pandemic as part of global
surveillance presents fundamental challenges to existing approaches
in two ways. Firstly, the sheer scale of data collected 
overwhelmed even approximate distance-matrix based methods.
% Accessed GISAID/EpiCov on 08 Oct., 2025
At the peak in February, 2022, about 80 thousand samples
were submitted on average per day to GISAID\cite{Shu2017},
accumulating to a total of over 17.5 million samples
as of October, 2025.
Regularly reinferring trees de-novo from such rapidly 
growing datasets is untenable.
The UShER\cite{Turakhia2021} method is based on adding 
samples to an existing tree using maximum parsimony, 
with periodic rearrangements to counter the inherent 
greediness of the approach\cite{DeMaio2023}.
The benefits of this simple model and the value of 
a comprehensive genealogy updated in real-time~\cite{McBroome2021}
were conclusively demonstrated, as the resource
quickly became an indispensable element 
of the pandemic response\cite{Hinrichs2024}.
The second challenge posed is that recombination is 
an important force in SARS-CoV-2 
evolution\cite{Jackson2021,VanInsberghe2021,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present in 
genealogy of the global population [cites??].
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{
Varabyou2021rapid,% Bolotie
Turakhia2022,%RIPPLES
Zhou2023virusrecom,% CovRecomb
Alfonsi2024,%RecombinHunt 
Li2024CovRecomb%CovRecomb 
},
this information is not incorporated into the phylogenies used 
in innumerable studies~\cite{McBroome2021,Hadfield2018},
and recombination continues to be treated as a 
nuisance factor.

@@

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts \cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango lineage naming system\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
In addition, we show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. By fully incorporating these and other novel
recombination events, the resulting data structure---representing
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
% Dunno about this! Feels important to finish on a wider note, helping
% the reader understand the significance of all of this?
These resources
will help us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

@@

\subsection*{Recombination intervals}
